---
title: "tools-graphs"
author: "Nikos Pechlivanis"
date: "2024-08-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load `libraries` 

```{r message=FALSE}
# install.packages("data.table")
library(data.table)
# install.packages("stringr")
library(stringr)

# install.packages(c("ggplot2", "ggrepel", "ggtext", "ggh4x"))
library(ggplot2)
library(ggrepel)
library(ggtext)
library(ggh4x)

# install.packages("extrafont")
library(extrafont)

library(paletteer)
library(colorspace)
```

## Load input dataset 

```{r loaddata, include = FALSE}

## THIS SECTION MIGHT THROUGH AN ERROR
## THERE IS A BUG UNFORTUNATELY WITH fread FUNCTION
## RUN ON CONSOLE IN THIS CASE

tools_url <- "https://raw.githubusercontent.com/galaxyproject/galaxy_codex/refs/heads/main/communities/microgalaxy/resources/curated_tools.tsv"

tools_dt <- tools_url |> fread()
```

## Tool Availability graph: Heatmap 

### Extract availability information

```{r}
index <- tools_dt |> colnames() |> str_subset("Number of tools on")

availability <- tools_dt[, c("Suite ID", "EDAM reduced topics", index), with = FALSE] |> unique()

colnames(availability) <- availability |> 
    colnames() |> 
    str_remove_all("Number\\ of\\ tools\\ on") |> 
    str_squish()
```

### Filter out duplicates

```{r}
index <- which(availability$`Suite ID` == "srst2" & availability$`EDAM reduced topics` == "")

availability <- availability[-index]
```

### Hierarchical clustering

```{r}
mm <- availability[, -c(1, 2)] |> as.matrix(rownames = availability$`Suite ID`)

mm_c <- mm |> dist(method = "manhattan") |> hclust(method = "ward.D2")
mm_r <- t(mm) |> dist(method = "manhattan") |> hclust(method = "ward.D2")
```

### Construct plotting data

```{r}
d <- availability |> melt(id.vars = c("Suite ID", "EDAM reduced topics"), variable.factor = FALSE, value.factor = FALSE)
d <- d[which(value >= 1)]

d$`Suite ID` <- d$`Suite ID` |> factor(levels = mm_c$labels[mm_c$order |> rev()])
d$variable <- d$variable |> factor(levels = mm_r$labels[mm_r$order |> rev()])

d$fct <- ifelse(d$variable |> str_detect("UseGalaxy"), "UseGalaxy", "vOther")

index <- d[which(fct == "UseGalaxy")][[1]] |> unique()

p <- d[which(`Suite ID` %in% index)]

p2 <- p |> tidyr::separate_rows("EDAM reduced topics", sep = ",") |> setDT()

p2$`EDAM reduced topics` <- p2$`EDAM reduced topics` |> str_squish()
p2$`EDAM reduced topics` <- p2$`EDAM reduced topics` |> str_wrap(width = 10)

t <- p2[, by = "EDAM reduced topics", .(N = `Suite ID` |> unique() |> length())]
t <- t[order(-N)]
t <- t[which(`EDAM reduced topics` != ""), head(.SD, 5)]

p2$edam_clean <- ifelse(p2$`EDAM reduced topics` %in% t$`EDAM reduced topics`, p2$`EDAM reduced topics`, "Other")
p2$edam_clean <- p2$edam_clean |> factor(levels = c(t$`EDAM reduced topics`, "Other"))
```

### Top 5 EDAM Operations

```{r}
c_1 <- p2[which(edam_clean != "Other")] |> 
    
    ggplot(aes(variable, `Suite ID`)) + 
    
    geom_tile(aes(fill = value), color = "grey") + 
    
    scale_fill_stepsn(
        colors = c('#00429d', '#5681b9', '#93c4d2', '#ffffe0', '#ffa59e', '#dd4c65', '#93003a'),
        guide = guide_colorsteps(barwidth = unit(14, "lines"), barheight = unit(.35, "lines")),
        breaks = c(2, 4, 8, 16, 32, 64, 128),
        transform = "log2"
    ) +
    
    facet_grid(cols = vars(fct), rows = vars(edam_clean), scales = "free", space = "free") +
    
    scale_y_discrete(expand = c(0, 0)) +
    scale_x_discrete(expand = c(0, 0)) +
    
    theme_minimal(base_family = "Calibri") +
    
    theme(
        legend.position = "bottom",
        legend.title.position = "top",
        
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 6),
        
        axis.title.x = element_markdown(),
        
        panel.border = element_rect(fill = NA, color = "grey25"),
        axis.ticks.x = element_line(lineend = "round", color = "grey25"),
        
        strip.text.x = element_blank(),
        strip.text.y = element_text(face = "bold", angle = 0, hjust = 0),
        
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dashed", lineend = "round", linewidth = .25),
        
        plot.margin = margin(10, 10, 10, 10)
    ) +
    
    labs(y = "Galaxy Tool Suites", x = "Availability of Tool Suites Across **Servers**", fill = "No. of Tools")
```

### All Tools suties

```{r}

c_2 <- p |> 
    
    ggplot(aes(`Suite ID`, variable)) + 
    
    geom_tile(aes(fill = value), color = "grey") + 
    
    scale_fill_stepsn(
        colors = c('#00429d', '#5681b9', '#93c4d2', '#ffffe0', '#ffa59e', '#dd4c65', '#93003a'),
        guide = guide_colorsteps(barwidth = unit(12, "lines"), barheight = unit(.5, "lines")),
        breaks = c(2, 4, 8, 16, 32, 64, 128),
        transform = "log2"
    ) +
    
    facet_grid(rows = vars(fct), scales = "free", space = "free") +
    
    scale_y_discrete(expand = c(0, 0)) +
    
    theme_minimal(base_family = "Calibri") +
    
    theme(
        legend.position = "bottom",
        legend.title.position = "top",
        
        axis.text.x = element_text(angle = 45, hjust = 1, size = 4),
        
        axis.title.y = element_markdown(),
        
        panel.border = element_rect(fill = NA, color = "grey10"),
        
        strip.text.y = element_blank(),
        strip.text.x = element_text(face = "bold", angle = 90, hjust = 0),
        
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dashed", lineend = "round", linewidth = .35)
    ) +
    
    labs(x = "Galaxy Tool Suites", y = "Availability of Tool Suites Across **Servers**", fill = "No. of Tools")

```

### Save plots

```{r}
outfolder <- "../results/tools/availability-heatmap"

dir.create(outfolder)

ggsave(
    plot = c_1, filename = paste0(outfolder, "/heatmap-1.png"),
    width = 5, height = 18, units = "in", dpi = 600
)

ggsave(
    plot = c_1, filename = paste0(outfolder, "/heatmap-1.svg"),
    width = 5, height = 18, units = "in"
)

ggsave(
    plot = c_1, filename = paste0(outfolder, "/heatmap-1.pdf"),
    width = 5, height = 18, units = "in", device = cairo_pdf
)

ggsave(
    plot = c_2, filename = paste0(outfolder, "/heatmap-2.png"),
    width = 5, height = 18, units = "in", dpi = 600
)

ggsave(
    plot = c_2, filename = paste0(outfolder, "/heatmap-2.svg"),
    width = 5, height = 18, units = "in"
)

ggsave(
    plot = c_2, filename = paste0(outfolder, "/heatmap-2.pdf"),
    width = 5, height = 18, units = "in", device = cairo_pdf
)
```


## EDAM operation: Scatter plot 

### Split EDAM operation per tool

```{r}
df2 = tools_dt |> tidyr::separate_rows("EDAM reduced operations", sep = ",") |> setDT()

xvar <- "Suite runs (last 5 years) on main servers"
yvar <- "Suite users (last 5 years) on main servers"
    
df2$runs  <- df2[[ xvar ]]
df2$users <- df2[[ yvar ]]

```

### Remove empty tools

```{r}
df2 = df2[which(runs != 0 & users != 0)]
```

### Clean `EDAM operation (no superclasses)` column

```{r}
df2$`EDAM reduced operations` = df2$`EDAM reduced operations` |> str_squish()
df2$`EDAM reduced operations` = ifelse(df2$`EDAM reduced operations` == "", "No Operation", df2$`EDAM reduced operations`)
```

### Find most top EDAM operations

```{r}
st = df2[, by = `EDAM reduced operations`, .(N = `Suite ID` |> unique() |> length())]
st = st[order(-N)]
st = st[which(`EDAM reduced operations` != "No Operation")]

df2$`EDAM reduced operations` <- df2$`EDAM reduced operations` |> factor(levels = c(st$`EDAM reduced operations`, "No Operation"))

df2 <- df2[order(`Suite ID`, -`EDAM reduced operations`)]
df2 <- df2[, by = `Suite ID`, head(.SD, 1)]
```

### Keep only necessary columns

```{r}
df2 = df2[, c("Suite ID", "runs", "users", "EDAM reduced operations")] |> unique()
```

### Define factor levels of cluster column

```{r}
st <- st[1:9]

df2$cluster <- ifelse(df2$`EDAM reduced operations` %in% st$`EDAM reduced operations`, df2$`EDAM reduced operations` |> as.character(), "Other")
df2$cluster <- df2$cluster |> factor(levels = c(st$`EDAM reduced operations`, "Other"))
```

### Define strip for plotting

```{r}
# df2$strip = paste0("**", df2$cluster, "** (", df2$N, " tools)")
# df2$strip = df2$strip |> factor(levels = df2$strip |> unique())
```

### gather highlighting Galaxy wrapper tools for plotting

```{r}
df2 <- df2[order(-runs, -users)]
hg0 <- df2[1:15]
```

### create graph

```{r}
ncolors <- df2$cluster |> unique() |> length()

xvar <- xvar |> str_replace("Suite runs", "**Suite runs**")
yvar <- yvar |> str_replace("Suite users", "**Suite users**")

a <- df2 |>
    
    ggplot(aes(runs, users)) +
    
    geom_point(
        shape = 21, size = 2, stroke = .25, 
        aes(fill = cluster, color = cluster)
    ) +
    
    geom_text_repel(
        data = hg0, aes(label = `Suite ID`),
        bg.r = .05, bg.color = "grey96", fontface = "bold", family = "Calibri", box.padding = .5,
        segment.size = .3, max.overlaps = Inf, size = 3
    ) +
    
    scale_x_continuous(
        trans = "log10",  # expand = c(0, 0), 
        limits = c(1, 10000000), 
        breaks = scales::trans_breaks("log10", function(x) 10^x),
        labels = scales::trans_format("log10", scales::math_format(10^.x))
    ) +
    
    scale_y_continuous(
        trans = "log10", # limits = c(1, 10000),
        labels = scales::comma, # expand = c(0, 0), 
        breaks = c(.1, 1, 10, 100, 1000, 10000)
    ) +
    
    scale_fill_manual(values = c(paletteer_d("ggsci::default_igv", ncolors - 1), "grey") |> lighten(.25), guide = guide_legend(nrow = 3, override.aes = list(size = 2.5))) +
    scale_color_manual(values = c(paletteer_d("ggsci::default_igv", ncolors - 1), "grey") |> darken(.25), guide = "none") +
    
    coord_cartesian() +
    
    theme_minimal(base_family = "Calibri") +
    
    theme(
        legend.position = "bottom",
        legend.title.position = "top",
        legend.justification = "left",
        
        strip.text = element_markdown(),
        
        axis.title.x = element_markdown(margin = margin(t = 10)),
        axis.title.y = element_markdown(margin = margin(r = 10)),
        
        # axis.ticks = element_line(linewidth = .3),
        panel.grid.major = element_line(linewidth = .3, linetype = "dashed", lineend = "round", color = "grey75"),
        
        # panel.border = element_rect(linewidth = .3, fill = NA),
        
        axis.line = element_line(lineend = "round"),
        axis.ticks = element_line(lineend = "round")
    ) +
    
    labs(x = xvar, y = yvar, fill = "EDAM operation")

# df3 <- df2[, by = .(`Galaxy wrapper id`, `Total tool usage (usegalaxy.eu)`, `No. of tool users (2022-2023) (usegalaxy.eu)`), .(
#     clusters = cluster |> sort() |> paste(collapse = ", "),
#     Nclusters = cluster |> length()
# )]
# 
# df3$clusters <- ifelse(
#     df3$clusters == "Other Operations" | df3$clusters == "Other Operations, No Operation",
#     "Other Operations", df3$clusters
# )
# 
# h1 <- h0[, c("Galaxy wrapper id", "Total tool usage (usegalaxy.eu)", "No. of tool users (2022-2023) (usegalaxy.eu)")] |>
#     unique()
# 
# a1 <- ggplot(df3, aes(`Total tool usage (usegalaxy.eu)`, `No. of tool users (2022-2023) (usegalaxy.eu)`)) +
#     
#     geom_point(shape = 21, size = 2, stroke = .2, color = "grey25", aes(fill = clusters)) +
#     
#     geom_text_repel(
#         data = h1, inherit.aes = FALSE,
#         mapping = aes(
#             `Total tool usage (usegalaxy.eu)`,
#             `No. of tool users (2022-2023) (usegalaxy.eu)`,
#             label = `Galaxy wrapper id`
#         ),
#         bg.color = "grey96", family = "Calibri", box.padding = .5,
#         segment.size = .3, max.overlaps = Inf
#     ) +
# 
#     scale_fill_manual(
#         values = paletteer_d("ggsci::default_igv"),
#         guide = guide_legend(ncol = 4)
#     ) +
#     
#     scale_x_continuous(
#         trans = "log10",  # expand = c(0, 0), 
#         limits = c(1, 10000000), 
#         breaks = scales::trans_breaks("log10", function(x) 10^x),
#         labels = scales::trans_format("log10", scales::math_format(10^.x))
#     ) +
#     
#     scale_y_continuous(
#         trans = "log10", # limits = c(1, 10000),
#         labels = scales::comma, # expand = c(0, 0), 
#         breaks = c(.1, 1, 10, 100, 1000, 10000)
#     ) +
#     
#     coord_cartesian() +
#     
#     theme_minimal(base_family = "Calibri") +
#     
#     theme(
#         legend.position = "bottom",
#         legend.justification = "left",
#         legend.title = element_blank(),
#         
#         strip.text = element_markdown(),
#         
#         axis.title.x = element_markdown(margin = margin(t = 10)),
#         axis.title.y = element_markdown(margin = margin(r = 10)),
#         
#         axis.ticks = element_line(linewidth = .3),
#         
#         panel.grid.minor = element_blank(),
#         panel.grid.major = element_line(linewidth = .3, linetype = "dashed", color = "grey75"),
#         
#         panel.border = element_rect(linewidth = .3, fill = NA),
#         
#         plot.margin = margin(20, 20, 20, 20)
#     ) +
#     
#     labs(
#         x = "**Total tool usage** (usegalaxy.eu)",
#         y = "**No. of tool users 2022-2023** (usegalaxy.eu)"
#     )

```

### Save plot

```{r}

outfolder <- paste0("../results/tools/", xvar |> str_remove_all("Suite|runs|\\*|\\(|\\)") |> str_squish())

dir.create(outfolder)

ggsave(
    plot = a, filename = paste0(outfolder, "/tools-scatter-plot.png"),
    width = 12, height = 9, units = "in", dpi = 600
)

ggsave(
    plot = a, filename = paste0(outfolder, "/tools-scatter-plot.svg"),
    width = 12, height = 9, units = "in"
)

ggsave(
    plot = a, filename = paste0(outfolder, "/tools-scatter-plot.pdf"),
    width = 12, height = 9, units = "in", device = cairo_pdf
)

```

## ARCHIVED EDAM topic: Faceted scatter plot 

```{r}
# # split EADM topics per tool ------------------------
# df2 = tools_dt |>
#     tidyr::separate_rows("EDAM topic (no superclasses)", sep = ",") |>
#     setDT()
# 
# # remove empty tools ------------------
# df2 = df2[which(
#     `Total tool usage (usegalaxy.eu)` != 0 & 
#         `No. of tool users (2022-2023) (usegalaxy.eu)` != 0
# )]
# 
# # clean `EDAM topic (no superclasses)` column -------------
# df2$`EDAM topic (no superclasses)` = df2$`EDAM topic (no superclasses)` |> str_squish()
# df2$`EDAM topic (no superclasses)` = ifelse(df2$`EDAM topic (no superclasses)` == "", "No Topic", df2$`EDAM topic (no superclasses)`)
# 
# # find most top #11 EDAM topic (no superclasses)s ------------------
# st = df2[, by = `EDAM topic (no superclasses)`, .(N = `Galaxy wrapper id` |> unique() |> length())]
# st = st[order(-N)]
# st = st[seq_len(11)]
# 
# # define every other operation as Other Operations ------------
# df2$cluster = ifelse(
#     df2$`EDAM topic (no superclasses)` %in% st$`EDAM topic (no superclasses)`,
#     df2$`EDAM topic (no superclasses)`, "Other Topics"
# )
# 
# # keep only necessary columns -------------------
# df2 = df2[, c(
#     "Galaxy wrapper id", 
#     "Total tool usage (usegalaxy.eu)", 
#     "No. of tool users (2022-2023) (usegalaxy.eu)", 
#     "cluster"
# )] |> unique()
# 
# # number of tools per operation ------------------------
# df2 = df2[, by = cluster, N := `Galaxy wrapper id` |> unique() |> length()]
# 
# # define factor levels of cluster column ----------------------
# df2$cluster = df2$cluster |> 
#     factor(
#         levels = c(
#             st$`EDAM topic (no superclasses)` |> str_subset("No Topic", negate = TRUE), 
#             "Other Topics", "No Topic"
#         )
#     )
# 
# df2 = df2[order(cluster)]
# 
# # define strip for plotting -------------------------------
# df2$strip = paste0("**", df2$cluster, "** (", df2$N, " tools)")
# df2$strip = df2$strip |> factor(levels = df2$strip |> unique())
# 
# # gather highlighting Galaxy wrapper tools for plotting ---------------
# 
# df2 <- df2[order(
#     -`No. of tool users (2022-2023) (usegalaxy.eu)`,
#     -`Total tool usage (usegalaxy.eu)`
# )]
# 
# h0 <- df2[, by = strip, head(.SD, 10)]
# 
# # h0 = rbind(
# #     df2[, by = strip, .SD[which.max(`Total tool usage (usegalaxy.eu)`)]],
# #     df2[, by = strip, .SD[which.max(`No. of tool users (2022-2023) (usegalaxy.eu)`)]]
# # ) |> unique()
# 
# # create graph ----------------------------
# b <- ggplot(df2, aes(`Total tool usage (usegalaxy.eu)`, `No. of tool users (2022-2023) (usegalaxy.eu)`)) +
#     
#     geom_point(shape = 21, size = 2, stroke = .25, 
#                color = "#2E2A2B" |> darken(.5), fill = "#2E2A2B" |> lighten(.5)) +
#     
#     geom_text_repel(
#         data = h0, inherit.aes = FALSE, 
#         mapping = aes(
#             `Total tool usage (usegalaxy.eu)`, 
#             `No. of tool users (2022-2023) (usegalaxy.eu)`, 
#             label = `Galaxy wrapper id`
#         ),
#         bg.color = "grey96", family = "Calibri", box.padding = .5,
#         segment.size = .3, size = 3, max.overlaps = Inf
#     ) +
#     
#     scale_x_continuous(
#         trans = "log10",  # expand = c(0, 0), limits = c(1, 10000000),
#         breaks = scales::trans_breaks("log10", function(x) 10^x),
#         labels = scales::trans_format("log10", scales::math_format(10^.x))
#     ) +
#     
#     scale_y_continuous(
#         trans = "log10", # limits = c(1, 10000),
#         labels = scales::comma, # expand = c(0, 0), 
#         breaks = c(.1, 1, 10, 100, 1000, 10000)
#     ) +
#     
#     guides(
#         alpha = guide_legend(
#             title = "Perc. of observations (tools)",
#             title.position = "top", 
#             title.theme = element_text(family = "Calibri")
#         )
#     ) +
#     
#     facet_wrap(vars(strip), nrow = 3, axes = "all") +
#     
#     coord_cartesian() +
#     
#     theme_minimal(base_family = "Calibri") +
#     
#     theme(
#         legend.position = "bottom",
#         legend.justification = "left",
#         
#         strip.text = element_markdown(),
#         
#         axis.title.x = element_markdown(margin = margin(t = 10)),
#         axis.title.y = element_markdown(margin = margin(r = 10)),
#         
#         axis.ticks = element_line(linewidth = .3),
#         
#         panel.grid.minor = element_blank(),
#         panel.grid.major = element_line(linewidth = .3, linetype = "dashed", lineend = "round", color = "grey75"),
#         
#         panel.border = element_rect(linewidth = .3, fill = NA)
#     ) +
#     
#     labs(
#         x = "**Total tool usage** (usegalaxy.eu)",
#         y = "**No. of tool users 2022-2023** (usegalaxy.eu)"
#     )

```


## No. of Tool Suites over time 

```{r}
d1 <- tools_dt[order(`Suite first commit date`)]

d1$ypos <- seq_along(d1$`Suite first commit date`)

d1$year <- d1$`Suite first commit date` |> lubridate::year()

d2 <- d1[, by = year, .(N = max(ypos))]

d2$year <- paste0(d2$year, "-01-01") |> lubridate::as_date()

d1_p <- d1 |>
    ggplot(aes(`Suite first commit date`, ypos)) +
    geom_line(linewidth = .45, color = "#3C3C3C") +
    geom_area(alpha = .25) +
    
    # geom_point(data = d2, aes(year, N)) +
    scale_x_date(expand = c(0.01, 0.01)) +
    scale_y_continuous(expand = c(0, 0), breaks = seq(50, 300, by = 50), limits = c(0, 235))+
    theme_minimal(base_family = "Calibri") +
    theme(
        # panel.grid.major = element_line(linewidth = .45, color = "grey85"),
        # panel.grid.minor = element_line(linewidth = .35),
        
        panel.grid.major = element_line(linewidth = .3, linetype = "dashed", lineend = "round", color = "grey75"),
        
        axis.title.x = element_markdown(margin = margin(t = 10)),
        axis.title.y = element_markdown(margin = margin(r = 10)),
        
        axis.line = element_line(lineend = "round"),
        axis.ticks = element_line(lineend = "round")
    ) +
    labs(y = "Cumulative number of tools")
```

### Save plot

```{r}

outfolder <- "../results/tools/number of tools"

dir.create(outfolder)

ggsave(
    plot = d1_p, filename = paste0(outfolder, "/ntools.png"),
    width = 9, height = 8, units = "in", dpi = 600
)

ggsave(
    plot = d1_p, filename = paste0(outfolder, "/ntools.svg"),
    width = 9, height = 8, units = "in"
)

ggsave(
    plot = d1_p, filename = paste0(outfolder, "/ntools.pdf"),
    width = 9, height = 8, units = "in", device = cairo_pdf
)
```


## Patchwork 

```{r}

library(patchwork)

multi <- (free(c_1) | (a / d1_p)) + 
    plot_layout(widths = c(2, 1.5)) +
    plot_annotation(tag_levels = "A") &
    theme(
        plot.tag = element_text(face = "bold", family = "Calibri", size = 14)
    )


outfolder <- "../results/tools/patchwork"
dir.create(outfolder)

ggsave(
    plot = multi, filename = paste0(outfolder, "/figure.png"),
    width = 14, height = 14, units = "in", dpi = 600
)

ggsave(
    plot = multi, filename = paste0(outfolder, "/figure.pdf"),
    width = 14, height = 14, units = "in", device = cairo_pdf
)

ggsave(
    plot = multi, filename = paste0(outfolder, "/figure.svg"),
    width = 14, height = 14, units = "in"
)

```